Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-03-08 article/Old versions/article_v1.tex
Redid folder structure + added first print/summary functions

@@

%\RepeatTitle{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}

@@

\subsection{Using PCA for data structure comparisons}
All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings. But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets. This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

@@

\subsubsection{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{21}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
%This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
%Anne: NK pointed out that this is somewhat confusing for people who are not familiar with the Bland-Altman plot, and as far as I can tell, what we do is quite different anyway, so I think no reference is really needed(?) 
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.

@@

\subsubsection{The angle plot}
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
%\begin{equation*}
%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
%= \lambda_{\max} \sum_{k=1}^d
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
%\end{equation*}
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand? Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions. The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

@@

\subsubsection{The chroma plot}
This plot compares the loadings of the two datasets. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segments are given by
\begin{align*}
\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
\end{align*}
where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively. Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
\begin{align*}
\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g. as identified by use of the angle plot. In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$.

@@

Moreover, though we have illustrated the discriminatory power of PCASDC in one data example, of course further evaluation of its performance is needed.  
While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness. For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible. Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots. We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots. First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.

@@

>>>>

\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes, positioning, calc}

@@

%\Large{\textsc{Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis}}

@@

\subsection{Using PCA for data structure comparisons}
\begin{figure}
\center 
Model 1: \medskip \\
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
  \node[sq node] (1) {$v_1$};
  \node[sq node, right of = 1] (2) {$v_2$};
  \node[sq node, right of = 2] (3) {$v_3$};
  \node[sq node, right of = 3] (4) {$v_4$};
  \node[sq node, right of = 4] (5) {$v_5$};
  \node[sq node, right of = 5] (6) {$v_6$};
  \node[main node, above right=1cm of 3] (7) {$X_1$};
  \node[main node, above right=1cm of 5] (8) {$X_2$};
  \path[every node/.style={font=\sffamily\small}]
    (7) edge node  {} (3)
    (7) edge node {} (4)
    (8) edge node {} (5)
    (8) edge node {} (6);
 \end{tikzpicture}
\\
\bigskip
Model 2: \medskip \\
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=2.5cm,
  thick,main node/.style={rectangle, rounded corners,draw,font=\sffamily\small\normalfont},
  sq node/.style={rectangle,draw,font=\sffamily\small\normalfont}]
  \node[sq node] (1) {$v_1$};
  \node[sq node, right of = 1] (2) {$v_2$};
  \node[sq node, right of = 2] (3) {$v_3$};
  \node[sq node, right of = 3] (4) {$v_4$};
  \node[sq node, right of = 4] (5) {$v_5$};
  \node[sq node, right of = 5] (6) {$v_6$};
  \node[main node, above=1cm of 2] (7) {$Y_1$};
  \node[main node, above=1cm of 5] (8) {$Y_2$};
  \path[every node/.style={font=\sffamily\small}]
    (7) edge node  {} (1)
    (7) edge node {} (2)
    (7) edge node {} (3)
    (8) edge node {} (4)
    (8) edge node {} (5)
    (8) edge node {} (6);
 \end{tikzpicture}
\caption{Graphical representations of the two models that are used for simulating example data for the PCADSC plots. Square nodes represent observed variables, while rounded nodes represent latent variables.  Arrows are used to illustrate causal links, and thus we e.g. find in \textit{Model 1} that the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable, namely $X_1$. Further details on the models and the simulated data are available in Appendix \ref{appendix.simData}, including covariance matrices for the two models.}
\label{fig.simGraphs}
\end{figure}

All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings. But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets. This can be the case while the two sets of scores could be arbitrarily different, which could e.g.\ happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

@@

While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure \ref{fig.simGraphs}. The observations are randomly divided into two groups.  
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure \ref{fig.simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description} 
Further details about the data simulations are available in Appendix \ref{appendix.simData}. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying datastructures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\subsubsection{The cumulative eigenvalue plot}
\begin{figure}
\centerline{\includegraphics[scale = 0.65]{simCE2.pdf}}
\centerline{\includegraphics[scale = 0.65]{simCE1.pdf}}
\caption{CE plots for the two simulated datasets. Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel. Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures. In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected). This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data. In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.}
\label{plot.simCE}
\end{figure}

The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{21}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
%This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
%Anne: NK pointed out that this is somewhat confusing for people who are not familiar with the Bland-Altman plot, and as far as I can tell, what we do is quite different anyway, so I think no reference is really needed(?) 
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.

@@

\subsubsection{The angle plot}
\begin{figure}
\centerline{\includegraphics[scale = 0.65]{simAngle2.pdf}}
\centerline{\includegraphics[scale = 0.65]{simAngle1.pdf}}
\caption{Angle plots for the two simulated datasets, with dataset A in the top panel and dataset B in the bottom panel. The blue show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse. For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.  For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
\label{plot.simAngle}
\end{figure}

This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
%\begin{equation*}
%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
%= \lambda_{\max} \sum_{k=1}^d
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
%\end{equation*}
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset %\hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand? Bo: Jeg har ikke noget bud på en sådan udvidelse lige nu og her --- skal vi ikke bare sige at det er ok i sin nuværende form?}.
We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions. The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length. Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.

@@

\subsubsection{The chroma plot}
\begin{figure}
\centerline{\includegraphics[scale = 0.65]{simChroma2.pdf}}
\centerline{\includegraphics[scale = 0.65]{simChroma1.pdf}}
\caption{Chroma plots for the two simulated datasets, with dataset A in the top and dataset B in the bottom. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance ($\tilde\sigma_{ij}^c$). For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model. For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.}
\label{plot.simChroma}
\end{figure}

This plot compares the loadings of the two datasets. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segments are given by
\begin{align*}
\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
\end{align*}
where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively. Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
\begin{align*}
\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g. as identified by use of the angle plot. In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$. In Figure \ref{plot.simChroma}, two examples of chroma plots are available.

@@

Moreover, though we have illustrated the discriminatory power of PCASDC in one simulated- and on real-life data example, of course further evaluation of its performance is needed.  
While the best feature of PCADSC is perhaps the intuitive, visual nature of the tools, it is also the biggest weakness. For statistical methods that result in a number (or a set of numbers), the standard way of evaluating performance is to report on simulation studies, but in the case of statistical methods that report graphics this is not feasible. Therefore, a systematic evaluation of the performance of the PCADSC methods is not very straight-forward to design: It really depends on whether or not users can detect actual differences in data structures by using the diagnostic plots. We suggest that the variability of the methods could be studied further using \textit{Wally plots} \citep{Ekstrøm2014}, which are learning tools developed for teaching students how to interpret residual plots. First in line as a topic for more thorough investigations of performance is the sensitivity towards the sample sizes of the two datasets, $n_1$ and $n_2$.

@@

\section{Details about the simulated data}
\label{appendix.simData}
In Section \ref{sec:pcadscintro}, we used simulated data for presenting the PCADSC plots. Dataset A consisted of 1000 independent simulated realizations from the $N(0, \Sigma_1)$-distribution, while dataset B consisted of 500 independent realizations from $N(0, \Sigma_1)$ and 500 independent realizations from $N(0, \Sigma_2)$. Dataset A was furthermore appended with a grouping variable, randomly dividing the observations into two groups. Dataset B also included a grouping variable, and this variable contained information about which of the two normal distributions each observation was simulated from. The covariance matrices, $\Sigma_1$ and $\Sigma_2$, were defined by

$$\Sigma_1 = \begin{pmatrix}
    1.0  &  0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
 0.0  &  1.0 & 0.0 & 0.0&  0.0 & 0.0 \\
 0.0   & 0.0 &  1.0 & 0.7 & 0.0 & 0.0 \\
 0.0 &  0.0 & 0.7 & 1.0 & 0.0 & 0.0 \\
0.0 &    0.0 &  0.0 & 0.0 & 1.0 & 0.4 \\
 0.0 & 0.0 &  0.0 & 0.0 & 0.4 & 1.0
\end{pmatrix}
\; \text{ and } \;
\Sigma_2 = \begin{pmatrix}
 1.0 & 0.2 & 0.1 & 0.0 & 0.0 & 0.0 \\
 0.2 & 1.0 & 0.1 & 0.0 & 0.0 & 0.0 \\
 0.1 & 0.1 & 1.0 & 0.0 & 0.0 & 0.0 \\
 0.0 & 0.0 & 0.0 & 1.0 & 0.3 & 0.1 \\
 0.0 & 0.0 & 0.0 & 0.3 & 1.0 & 0.2 \\
 0.0 & 0.0 & 0.0 & 0.1 & 0.2 & 1.0
\end{pmatrix}
$$,

respectively.
